% ---------------------------------------------------------------------------
% MLSys 2027 — main-track submission (skeleton, A-stage of A→B staging)
%
% Status: skeleton + section outlines + framing prose. Body to be filled
% in iteratively. Empirical numbers in §5 (Evaluation) are marked
% [TODO:bench] and depend on the B-stage bench harness landing
% (paper/mlsys-2027/README.md → "A → B staging").
%
% Build: `make` in this directory (pdflatex + bibtex + pdflatex × 2).
% Venue style file (mlsys2027.sty) will be swapped in when the CFP
% drops; until then we compile with the generic article class.
% ---------------------------------------------------------------------------

\documentclass[11pt]{article}

% --- Geometry / typography ---
\usepackage[letterpaper,margin=1in]{geometry}
\usepackage{microtype}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{times}

% --- Math ---
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{mathtools}

% --- Tables / figures ---
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{subcaption}

% --- Code listings ---
\usepackage{listings}
\usepackage{xcolor}
\lstset{
  basicstyle=\ttfamily\small,
  commentstyle=\color{gray!70!black}\itshape,
  keywordstyle=\color{blue!50!black}\bfseries,
  stringstyle=\color{red!60!black},
  showstringspaces=false,
  breaklines=true,
  numbers=left,
  numberstyle=\tiny\color{gray},
  frame=single,
  framerule=0pt,
  backgroundcolor=\color{gray!8},
  xleftmargin=2em,
  xrightmargin=1em,
  belowskip=-0.5em
}
\lstdefinelanguage{cppx}[]{C++}{
  morekeywords={requires,concept,consteval,constinit,co_await,co_return,co_yield},
  moredelim=[s][\color{purple!70!black}]{"}{"_tex},
}

% --- References / hyperlinks ---
\usepackage[hidelinks]{hyperref}
\usepackage{cleveref}
\usepackage{url}

% --- TODO marker (no external package dep) ---
\newcommand{\TODO}[1]{\textbf{\textcolor{red!70!black}{[TODO: #1]}}}
\newcommand{\BENCH}[1]{\textbf{\textcolor{orange!70!black}{[BENCH: #1]}}}

% --- Domain macros (will migrate to paper/common/symbols.tex) ---
\newcommand{\NamedTensor}[1]{\mathsf{#1}}
\newcommand{\ax}[1]{\textsf{#1}}                              % axis name
\newcommand{\udltex}{\code{\_tex}}                            % _tex UDL (argless)
\newcommand{\code}[1]{\texttt{#1}}

% ---------------------------------------------------------------------------

\title{Named Axes as a Systems Discipline:\\
       Hybrid Compile-Time and Runtime Coordination\\
       for Tensor Programs}

\author{%
  \TODO{author block — fill at submission time}\\
  \texttt{authors@example.com}
}

\date{}

\begin{document}
\maketitle

% ===========================================================================
\begin{abstract}
Tensor programs routinely fail on \emph{axis-bug} classes that
positional shape systems cannot catch: silent broadcasts across
mis-named dimensions, transposed contractions, reductions over the
wrong axis. Existing named-axis libraries push the check from
``silent wrong output'' to ``runtime exception'' but leave the host
language's static type system unused. We present a hybrid design that
encodes axis information across the compile-time / runtime boundary,
with three contributions: (i) a \udltex{} user-defined literal that
parses a LaTeX-math subset (e.g.\
\code{R"(c\_\{ij\} = a\_i b\_j)"\_tex}) into a named-axis expression
AST whose structural indices the type system can reason about; (ii) a
hybrid \emph{NTTP $\times$ DynamicShape} axis system that refuses
label mismatches at template instantiation when shapes are known and
falls back to runtime guards otherwise --- one kernel implementation
across both paths (\Cref{app:static-catch-demo} records a verbatim
\code{static\_assert} refusal as the minimum-working-example); and
(iii) a \emph{Hexagonal-lite} kernel-backend port carrying three
substrates (reference, Eigen, WebGPU/Dawn) behind one C++20
\code{concept}. Our axis-bug benchmark suite and a bundle-adjustment
case study on the ETH3D dataset operationalise the catch-rate and
substrate-cost claims.
\end{abstract}

% ===========================================================================
\section{Introduction}
\label{sec:intro}

% ~1 page. Lead with the failure-mode story (axis bug), position
% against the landscape, state contributions, lay out roadmap.

Consider a researcher writing a multi-head attention block. They
compose two linear projections, a softmax, and a matmul. The shapes
type-check at every step --- every intermediate has the right number of
dimensions --- yet the network silently produces incorrect outputs at
training time, because two axes were transposed five lines up. The
type system saw a $(B,H,T,D)$ tensor where the writer meant
$(B,T,H,D)$; the names lived only in a comment.

This failure mode is endemic. PyTorch's named-tensor prototype was
introduced in 2019 specifically to refute it; six years on it remains
labelled ``prototype, subject to change'' and the
community-tracking issue records that development has
stalled~\cite{pytorch-named-tensors}.
JAX's analogous \code{xmap} was deleted in 2024~\cite{jax-xmap-deletion}.
The bug is well-known; what is missing is a design that the host
language's type system can refuse, not merely a runtime that throws.
Existing libraries respond along two axes:

\begin{itemize}
  \item \textbf{String-based runtime notation} (\code{einops}~\cite{rogozhnikov2022einops},
        \code{einx}~\cite{flick2026einx}): expressive (Einstein-style formulas
        are the source of truth) but defers all checking to runtime
        and forfeits the host language's type system.
  \item \textbf{Language-level named tensors with runtime checks}
        (\code{Haliax}~\cite{haliax}, \code{Penzai}~\cite{penzai},
        \code{xarray}~\cite{xarray}, deprecated
        \code{torch.named\_tensor}~\cite{pytorch-named-tensors}):
        axis names are first-class values, but verification still
        happens at runtime.
\end{itemize}

\paragraph{The thesis.}
A named-axis library should be able to refute the attention-block bug
\emph{at compile time, when the shapes are known}, while still
permitting the runtime-shape case that batch processing demands. The
test of a design is not whether axis names are present but whether they
participate in the type system.

\paragraph{Contributions.}
This paper presents the \code{tensor} library, organised around three
design moves:

\begin{enumerate}
  \item \textbf{The \udltex{} LaTeX bridge} (\Cref{sec:udl}): a
        user-defined literal that parses a LaTeX-math subset
        (\code{R"(c\_\{ij\} = a\_i b\_j)"\_tex}) into a named-axis
        \code{Expression} AST. The AST's index structure --- the
        names attached to each \code{IndexedVar} and the bound index
        on each \code{Sum} --- is what the type system reasons about,
        so the same formula composes with the static-axis machinery
        of contribution~(ii) at no boundary cost.
  \item \textbf{Hybrid axis system} (\Cref{sec:hybrid}): non-type
        template parameters carry compile-time axis sizes;
        \code{DynamicShape} carries the runtime-determined cases; the
        two compose without bifurcating the kernel API.
  \item \textbf{Hexagonal-lite kernel-backend port}
        (\Cref{sec:hex-lite}): a single port boundary admits three
        substrates (reference, Eigen, WebGPU-via-Dawn~\cite{dawn})
        without leaking substrate concerns into the named-axis layer.
\end{enumerate}

\paragraph{Roadmap.}
\Cref{sec:bg} surveys the axis-bug taxonomy and the prior-art landscape.
\Cref{sec:udl,sec:hybrid,sec:hex-lite} present the three design moves.
\Cref{sec:eval} reports the static-catch benchmark and a
bundle-adjustment case study across three backends. \Cref{sec:related}
positions against contemporaneous work. \Cref{sec:discussion} discusses
limitations.

% ===========================================================================
\section{Background \& Motivation}
\label{sec:bg}

% ~1 page. Three subsections: axis-bug taxonomy, prior-art landscape
% (compressed from M1 study), the open gap our design fills.

\subsection{The Axis-Bug Taxonomy}
\label{sec:bg-taxonomy}

We catalogue five recurring axis-bug classes (\Cref{tab:axis-bugs}).
Each class is shape-compatible --- the offending program type-checks,
runs, and silently produces incorrect output --- so a position-typed
shape system cannot refute any of them.

\begin{table*}[t]
\centering\small
\begin{tabular}{@{}p{0.18\linewidth} p{0.42\linewidth} p{0.32\linewidth}@{}}
\toprule
\textbf{Class} & \textbf{One-line reproduction} & \textbf{What is silently wrong} \\
\midrule
\textbf{T1.} Transposed contraction &
\code{attn = Q @ K.transpose(-2,-1)} where \code{Q} is laid out
\code{(B,T,H,D)} rather than the expected \code{(B,H,T,D)} &
Contraction sums over the head axis, not over the model dim.
Scores have the right shape, wrong meaning. \\
\addlinespace[2pt]
\textbf{T2.} Silent broadcast &
\code{x + bias.T} where \code{x} is \code{(32,128)} and \code{bias}
is \code{(1,128)} &
The transposed bias \code{(128,1)} stretches against the batch axis
instead of the feature axis; every row gets a different bias scalar. \\
\addlinespace[2pt]
\textbf{T3.} Reduction over wrong axis &
\code{loss = -(targets * x.log()).sum(dim=0)} on logits
\code{x: (B, C)} &
Reduces over the batch dimension; the resulting loss vector has
length $C$ instead of being a scalar. Loss curve looks
``surprisingly small'' and training silently diverges. \\
\addlinespace[2pt]
\textbf{T4.} Layout swap &
A module documented as returning \code{(H, B, D)} is consumed by code
expecting \code{(B, H, D)} and a \code{.view(B, H, D)} is applied
without a \code{.permute(1,0,2)} first &
Adjacent batch entries get mixed across heads; gradients flow
incorrectly. The most common production failure mode (PyTorch named
tensors~\cite{pytorch-named-tensors} were originally introduced to
address exactly this). \\
\addlinespace[2pt]
\textbf{T5.} Stack vs.\ concat &
\code{torch.cat([a, b], dim=0)} where the intent was to add a new
length-2 axis &
Produces a tensor of rank $r$ rather than rank $r{+}1$; downstream
\code{.view} calls then either silently misalign or crash several
operators later, far from the original bug site. \\
\bottomrule
\end{tabular}
\caption{Axis-bug taxonomy. All five classes are shape-compatible:
no positional shape system rejects them. \Cref{sec:eval-static}
benchmarks each library's static / runtime / silent rate against an
expanded version of this taxonomy.}
\label{tab:axis-bugs}
\end{table*}

\subsection{The Landscape}
\label{sec:bg-landscape}

Existing approaches to axis naming differ on two orthogonal axes:
\emph{where the axis name lives} and \emph{when it is checked}
(\Cref{tab:landscape}). The space splits into three clusters:

\begin{table}[t]
\centering\small
\begin{tabular}{@{}p{0.30\linewidth} p{0.30\linewidth} p{0.30\linewidth}@{}}
\toprule
\textbf{Library} & \textbf{Carrier} & \textbf{Check phase} \\
\midrule
\code{einops}~\cite{rogozhnikov2022einops}     & string literal              & runtime \\
\code{einx}~\cite{flick2026einx}         & string literal              & runtime \\
\code{xarray}~\cite{xarray}     & runtime value (attribute)   & runtime \\
\code{Haliax}~\cite{haliax}     & runtime value (\code{NamedArray}) & JAX trace-time$^\dagger$ \\
\code{Penzai}~\cite{penzai}     & runtime value (\code{NamedArray} PyTree) & JAX trace-time$^\dagger$ \\
\code{torch.named\_tensor}~\cite{pytorch-named-tensors}
                                & runtime value (\code{names=})
                                                              & runtime (stalled) \\
\code{Nx}~\cite{nx-elixir}      & first-class (Elixir)        & dialyzer-time (partial) \\
\code{tinygrad}~\cite{tinygrad} & none (positional)           & --- \\
\midrule
\code{tensor} (this work)       & NTTP (\code{LabelTag<"i">}) \emph{or}
                                  runtime \code{Axis}, per tensor type
                                & \textbf{compile time when shapes are
                                  static; runtime via the Evaluator
                                  otherwise} \\
\bottomrule
\end{tabular}
\caption{Where axis names live, and when they are checked.
$^\dagger$\emph{JAX trace-time} is the shape check that fires at the
first \code{jit}/\code{pjit} call --- meaningfully later in the dev
loop than C++ template instantiation but still before runtime
proper. The \code{tensor} library uniquely occupies the
template-instantiation-time-when-known quadrant via
\code{LabelTag<"i">} NTTPs, with explicit fallback to runtime via
\code{DynamicShape}.}
\label{tab:landscape}
\end{table}

\paragraph{String-based runtime.}
\code{einops}~\cite{rogozhnikov2022einops} and \code{einx}~\cite{flick2026einx} encode axis
intent in formula strings interpreted on each call. The notation is
expressive (Einstein-style patterns drive contractions, reshapes, and
reductions from one source-of-truth string) but every check fires at
runtime and nothing about the formula reaches the host language's
type system.

\paragraph{Named-value runtime.}
\code{Haliax}~\cite{haliax}, \code{Penzai}~\cite{penzai},
\code{xarray}~\cite{xarray}, and the prototype \code{torch.named\_tensor}
\cite{pytorch-named-tensors} carry axis names as runtime values
attached to each tensor (a \code{NamedArray} field, an
\code{xarray.DataArray} dimension attribute, the \code{names=} kwarg).
Axis mismatch is caught at the first operation that touches both
tensors --- the failure mode is no longer ``silent wrong output'' but
``crash at runtime,'' which is strictly better, yet still misses the
opportunity to refuse the program at compile time when the shapes
are known. The PyTorch implementation has remained labelled
``prototype'' since its 2019 introduction and development has
stalled~\cite{pytorch-named-tensors}; JAX's analogous \code{xmap} was
deleted in 2024 in favour of \code{shard\_map}~\cite{jax-xmap-deletion}.

\paragraph{Type-level (rare).}
Outside of compiler-IR frameworks (TACO, MLIR \code{linalg}), only
small static-C++ entrants attempt type-level axis names: Einsums
encodes index packs as templated literals, Fastor encodes positions as
template parameters. Neither admits the runtime-shape case without
escape-hatching back to untyped views.

\subsection{What remains open}
\label{sec:bg-gap}

The landscape leaves an explicit gap: \emph{compile-time axis checking
that does not abandon the runtime-shape case}. The static approaches
in the literature require either fully static shapes (no batch
flexibility) or full type erasure at the kernel boundary. Our design
target is to recover the latter without giving up the former.

% ===========================================================================
\section{Design 1: The \texorpdfstring{\udltex{}}{\_tex} LaTeX Bridge}
\label{sec:udl}

% ~1.5 pages. The honest framing: today's UDL is a runtime LaTeX-subset
% parser; its contribution is structural (the AST shape composes with
% the static-axis type system) rather than temporal (when the parse
% fires). The scheduled consteval lift is § Future Work, not here.

\subsection{Interface}
\label{sec:udl-interface}

The user writes the formula the way the corresponding paper would, in
a C++ raw-string literal suffixed \code{\_tex}, and feeds it through
an Evaluator bound to named tensors:

\begin{lstlisting}[language=cppx]
// Bind names -> tensors, then evaluate the formula.
tex::Evaluator<double> ev;
ev.bind("a", DynamicTensor{Shape{Axis{"i", 5}}, /*data*/});
ev.bind("b", DynamicTensor{Shape{Axis{"j", 5}}, /*data*/});

auto c = ev.evaluate(R"(c_{ij} = a_i b_j)"_tex);
// c has named axes (i, j); c[i,j] == a[i] * b[j].
\end{lstlisting}

The literal \code{R"(c\_\{ij\} = a\_i b\_j)"\_tex} parses on
construction into an \code{Expression} AST. The
\code{tex::Evaluator} walks the AST against name-to-tensor bindings,
asserting that each \code{IndexedVar}'s subscript list matches its
bound tensor's axis labels in order. For example,
\code{R"(a\_\{ji\})"\_tex} bound against a tensor whose axes are
declared $(\ax{i},\ax{j})$ is rejected by the Evaluator's
subscript-order check.

\subsection{Grammar}
\label{sec:udl-grammar}

The accepted grammar is the LaTeX-math subset useful for naming and
combining axis-indexed quantities (informal BNF in priority order):

\begin{lstlisting}[basicstyle=\ttfamily\footnotesize,numbers=none]
equation    := expression ('=' expression)?
expression  := term (('+' | '-') term)*
term        := factor (('*' | '\cdot' | juxt) factor)*
factor      := ('-' factor) | unary
unary       := atom (('/' atom)*)
atom        := '(' expression ')'
             | '\sum' '_' subscript group
             | name subscript?
subscript   := '_' (letter | '{' letters '}')
group       := '{' expression '}' | atom
\end{lstlisting}

Whitespace is ignored; juxtaposition is implicit multiplication. The
parser is a hand-written recursive descent; the grammar is small
enough that the implementation fits in a single header
(\code{include/tensor/tex/parser.hpp}).

The choice of LaTeX over an einops-style mini-language is deliberate.
The subset that matters --- subscripted variables, Einstein sums
(\code{R"(\textbackslash sum\_i \{a\_i b\_i\})"\_tex}), and equations
that name a result --- is the notation tensor-program papers and
textbooks already use. The literal is also \emph{readable as LaTeX},
so the same text that drives evaluation can be re-emitted into the
program's typeset output: a Jupyter cell prints back the original
formula, closing the loop from notation to running code to rendered
documentation.

\subsection{Composability with the Static-Axis System}
\label{sec:udl-composability}

The system-level claim of this section is not about \emph{when} the
parse fires --- it fires at literal construction today, which is
runtime --- but about \emph{what shape} it produces. The parsed AST's
indices are the same structural objects the static-axis machinery of
\Cref{sec:hybrid} reasons about. The Evaluator's binding step is a
type-level operation when the bound tensor's axes are NTTP-encoded,
and a runtime check when they are carried in a \code{DynamicShape}.
The Evaluator does not branch on which world it is in; one code path
services both.

Concretely, the same literal
\code{R"(c\_\{ij\} = a\_i b\_j)"\_tex} drives three different
compile-time / runtime mixes without changing form:

\begin{lstlisting}[language=cppx]
using namespace tensor::core::literals;       // for "i"_ax, "j"_ax
constexpr auto i = "i"_ax;
constexpr auto j = "j"_ax;

// Scenario (a): both operands NTTP-labelled.
TypedTensor<double, "i"_ax> a_t{{5}, {/*...*/}};
TypedTensor<double, "j"_ax> b_t{{5}, {/*...*/}};
// Bind via to_dynamic(); the label match is a type-level operation
// at template instantiation, sizes are propagated by construction.

// Scenario (b): one NTTP, one DynamicShape (e.g. batch dim from disk).
TypedTensor<double, "i"_ax> a_t2{{5}, {/*...*/}};
DynamicTensor<double> b_d{Shape{Axis{"j", k}}, /*data*/};
// At ev.bind("b", ...): the name 'j' is checked at the call site;
// the runtime extent k is propagated into the Expression's shape.

// Scenario (c): both DynamicShape (e.g. inference on serialised tensors).
DynamicTensor<double> a_d{Shape{Axis{"i", n}}, /*data*/};
DynamicTensor<double> b_d2{Shape{Axis{"j", m}}, /*data*/};
// Both name and size are runtime; both checks fire at ev.bind() and
// at ev.evaluate(...) time. Failure is a ShapeError, not undefined
// behaviour.

tex::Evaluator<double> ev;
ev.bind("a", /* a_t.to_dynamic() or a_d */);
ev.bind("b", /* b_t.to_dynamic() or b_d  */);
auto c = ev.evaluate(R"(c_{ij} = a_i b_j)"_tex);
\end{lstlisting}

The literal text is identical across (a)--(c). The Evaluator's
implementation is identical across (a)--(c). What changes is only
\emph{at which moment} each label and each size is verified: at type
instantiation, at \code{bind} time, or at \code{evaluate} time. The
hybrid system pays for the strongest verification available given
how much was known at each construction site --- and no more.

\subsection{Scope and the Scheduled \texorpdfstring{\code{consteval}}{consteval} Lift}
\label{sec:udl-limits-and-future}

The parser is presently runtime, executed at the literal's
construction site. The natural \code{consteval} formulation is
blocked by C++20's prohibition on \code{std::unique\_ptr} escaping a
constant-evaluated context (the current AST is a
\code{unique\_ptr}-of-variant tree). A scheduled refactor replaces
the \code{unique\_ptr} representation with a fixed-size
arena-of-variants whose lifetime is the literal's; that representation
does escape constant evaluation and lifts the entire parse into the
compiler.

Three properties of the current design are unchanged by the lift, and
are this section's contribution:

\begin{enumerate}
  \item the parser's output type (\code{Expression}) is identical in
        both worlds;
  \item the Evaluator's binding-and-check pass is identical;
  \item the parse cost is paid once per literal, not once per call
        --- today at static-initialiser time, after the lift at
        compile time.
\end{enumerate}

Grammar-wise, the deliberate non-goals are: no ellipsis, no broadcast
operators, no rank-polymorphic patterns, no run-time--constructed
strings (the literal arrives as \code{char const*}). Cases the
grammar cannot express are carried by the hybrid system of
\Cref{sec:hybrid}.

% ===========================================================================
\section{Design 2: Hybrid NTTP $\times$ DynamicShape}
\label{sec:hybrid}

% ~1.5 pages. Two compile-time paths and one runtime path; how they
% compose at the kernel boundary.

\subsection{The two paths}

A named tensor carries its axis information in one of two ways:

\begin{description}
  \item[Compile-time path] axis names \emph{and} sizes are encoded as
        non-type template parameters. A literal expecting axes
        $(\ax{i},\ax{j})$ --- e.g.\
        \code{R"(c\_\{ij\} = a\_i b\_j)"\_tex} --- bound through the
        Evaluator against a tensor whose NTTP axes are
        $(\ax{i},\ax{k})$ is refuted at template instantiation; the
        diagnostic names the offending axis.
  \item[Runtime path] axis names are encoded as template parameters
        (compile-time) but sizes live in a \code{DynamicShape} value
        (runtime). The compiler refutes axis-\emph{name} mismatches;
        size mismatches surface as a \code{ShapeError} at the kernel
        call site.
\end{description}

\subsection{Promotion and Demotion}
\label{sec:hybrid-promo}

The two paths are explicitly bridged by one demotion entry point
(\code{TypedTensor::to\_dynamic()}) and no implicit promotions.

\paragraph{Demotion (CT~$\to$~RT) is total and free.}
A \code{TypedTensor<T, "i"\_ax, "j"\_ax>} carries its labels as
\code{FixedString} NTTPs and its extents as a static-rank
\code{Shape<2>}. Its \code{to\_dynamic()} method copies the underlying
buffer into a \code{DynamicTensor<T>} whose \code{DynamicShape} is
populated by reading the NTTP labels out at compile time and the
extents at runtime --- the names \emph{survive the demotion as
runtime strings}. The conversion is total: every TypedTensor demotes;
nothing is lost except the type-level guarantee.

\begin{lstlisting}[language=cppx]
TypedTensor<double, "i"_ax, "j"_ax> t{{2, 3}, /*values*/};
DynamicTensor<double> d = t.to_dynamic();
// d.shape() now has Axis{"i", 2}, Axis{"j", 3}.
// The type information is erased; the *name* information is not.
\end{lstlisting}

This is what lets a generic Evaluator-driven kernel
(\Cref{sec:udl}) accept TypedTensor inputs without requiring a
templated overload per static-axis pack: the kernel sees one
\code{DynamicTensor} type and the name check uses runtime strings,
which the demoted tensor still carries.

\paragraph{Promotion (RT~$\to$~CT) is opt-in and checked.}
Going the other direction is not type-safe by construction --- the
labels read from a \code{DynamicShape} are runtime data --- so the
library does not provide an implicit conversion. Instead, the user
asserts the expected static labels and the runtime data is verified
once:

\begin{lstlisting}[language=cppx]
DynamicTensor<double> d = load_from_disk(...);
// d.shape() carries runtime Axis labels and extents.

// Promotion asserts the expected NTTP pack; if the labels do not
// match (or rank differs), construction throws std::invalid_argument
// at the assertion site rather than silently producing a wrong type.
auto t = TypedTensor<double, "i"_ax, "j"_ax>::from_dynamic(d);
\end{lstlisting}

This is the only point in the system where the runtime label string
is read into a type-level decision. The choice to make promotion
explicit means the type system retains a useful invariant: a
\code{TypedTensor<T, L...>} value's labels are always
\code{L...}~--- always known at compile time, never doubted at
runtime.

\paragraph{The two-path rule.}
A kernel is written exactly once, against the runtime path. Static
callers reach it via \code{to\_dynamic()}; dynamic callers reach it
directly. The compiler-enforced \code{static\_assert(SameLabels<A,B>)}
guard in the \code{TypedTensor} operators (\code{operator+},
\code{operator-}, etc.) refuses static-side mismatches before they
reach the demotion call; the Evaluator's name-and-rank checks refuse
the same mismatches on the dynamic side. The two checks share the
intent (``the labels do not match'') and differ only in when they
fire.

\subsection{Kernel boundary}

The contract of a kernel is a function template parameterised by the
axis structure (compile-time) accepting \code{DynamicShape} runtime
sizes when needed. \emph{The same kernel} services both paths; the
runtime-path overhead is one bounds check per call, not a separate
kernel implementation.

% ===========================================================================
\section{Design 3: Hexagonal-lite Kernel Backend}
\label{sec:hex-lite}

% ~1 page. The port; the three adapters; why this matters for MLSys
% specifically (substrate evolution, e.g., webgpu, is the active
% research area).

The named-axis layer should not know whether contractions are computed
by hand-rolled loops, an Eigen template instantiation, or a WGSL
kernel dispatched to a GPU. \emph{Hexagonal-lite} is a constrained
form of Cockburn's Hexagonal architecture~\cite{cockburn-hexagonal}: a single port
(\code{KernelBackend}) with three concrete adapters.

\subsection{The Port}
\label{sec:hex-port}

The port is a C++20 \code{concept} declared in
\code{include/tensor/core/concepts.hpp}. A substrate that satisfies
\code{KernelBackend} can be plugged into the Domain without further
adaptation work:

\begin{lstlisting}[language=cppx]
template <class B>
concept KernelBackend = requires(B b) {
    typename B::backend_tag;
} && requires(B b,
              DynamicTensor<double> const& a,
              DynamicTensor<double> const& a2,
              BroadcastPlan const& bplan,
              ContractPlan const& cplan,
              std::vector<std::size_t> const& source_map,
              DynamicShape const& source_shape) {
    // Element-wise binary, same shape.
    { b.add(a, a2) } -> std::same_as<DynamicTensor<double>>;
    { b.sub(a, a2) } -> std::same_as<DynamicTensor<double>>;
    { b.mul(a, a2) } -> std::same_as<DynamicTensor<double>>;
    { b.div(a, a2) } -> std::same_as<DynamicTensor<double>>;
    // Element-wise unary.
    { b.exp(a)  } -> std::same_as<DynamicTensor<double>>;
    { b.log(a)  } -> std::same_as<DynamicTensor<double>>;
    { b.relu(a) } -> std::same_as<DynamicTensor<double>>;
    { b.neg(a)  } -> std::same_as<DynamicTensor<double>>;
    // Broadcast element-wise.
    { b.broadcast_add(a, a2, bplan) } -> std::same_as<DynamicTensor<double>>;
    { b.broadcast_sub(a, a2, bplan) } -> std::same_as<DynamicTensor<double>>;
    { b.broadcast_mul(a, a2, bplan) } -> std::same_as<DynamicTensor<double>>;
    // Contraction (Einstein-style).
    { b.contract(a, a2, cplan) } -> std::same_as<DynamicTensor<double>>;
    // Reduction.
    { b.reduce_sum(a) } -> std::same_as<double>;
    // Unbroadcast: used by autograd to reduce dL/dout to an input's shape.
    { b.unbroadcast(a, source_map, source_shape) }
        -> std::same_as<DynamicTensor<double>>;
};
\end{lstlisting}

Three deliberate choices shape this surface:

\begin{enumerate}
  \item \textbf{The Domain pre-computes the plans.} \code{BroadcastPlan}
        and \code{ContractPlan} are computed once in
        \code{tensor::core} and handed to the backend, so a substrate
        never re-implements axis matching or reduction-set inference.
        Substrates differ on \emph{how} to evaluate a plan, not on
        \emph{whether} a plan is well-formed.
  \item \textbf{Unbroadcast is in the port.} The reverse of a
        broadcasted op --- reducing \code{dL/dout} back to an input's
        shape --- belongs to the substrate because Eigen and Kokkos
        can implement it as a single fused reduction. Keeping it
        outside autograd lets the autograd layer compose port methods
        without per-substrate special cases.
  \item \textbf{Per-op backward closures are \emph{not} in the port.}
        They live in \code{tensor::autograd} and compose port
        methods. The substrate owns ``how to compute,'' the autograd
        layer owns ``how to differentiate.''
\end{enumerate}

Each concrete substrate ends with a
\code{static\_assert(KernelBackend<Backend>)}, so any divergence
between the port surface and an adapter is a compile error at the
point where the adapter is defined, not a runtime surprise.

\subsection{Substrates}

\begin{description}
  \item[Reference] portable naive nested-loop implementation; defines
        the semantics that the other substrates must match.
  \item[Eigen] BLAS-routed contractions for CPU performance baseline;
        loop fusion via Eigen's expression templates.
  \item[WebGPU via Dawn]~\cite{dawn} GPU substrate compiled to
        WGSL; vcpkg-vendored Dawn to keep substrate provisioning
        reproducible in CI.
\end{description}

The choice of substrate is a CMake variable
(\code{-DTENSOR\_KERNEL\_BACKEND=\{reference,eigen,webgpu\}}) ---
the named-axis layer is unchanged across all three.

% ===========================================================================
\section{Evaluation}
\label{sec:eval}

We evaluate two distinct questions: \emph{does the static-axis path
actually refuse the axis-bug classes of \Cref{sec:bg-taxonomy}?}
(\Cref{sec:eval-static}), and \emph{does the Hexagonal-lite
substrate boundary leak performance through its abstraction?}
(\Cref{sec:eval-ba}). The static-catch numbers are the paper's
distinctive evaluation; the runtime case study sanity-checks that
substrate substitutability does not cost an order of magnitude.

\paragraph{Status of results.}
Methodology, harness scaffolding, and the comparison protocol are
final. Numerical results are collected by a dedicated benchmark
harness (\code{bench/bench\_mlsys\_paper\_case\_study.py}),
implementation of which is scheduled for September 2026, ahead of
the conference deadline; the present submission carries the
fully-specified protocols (this section) and an empty result table
that the harness output drops into directly. Where numbers below
read \BENCH{$\cdot$}, the design of the measurement is locked but
the number itself is pending the harness landing.

\subsection{Static-catch Benchmark}
\label{sec:eval-static}

\paragraph{Suite construction.}
The benchmark suite has $5 \times 10 = 50$ \emph{bug instances}
(ten per axis-bug class T1--T5 of \Cref{tab:axis-bugs}) plus 25
\emph{control programs} that exercise the same operators
\emph{correctly}. The controls let us also report a per-library
\emph{false-positive rate}: a system that refuses everything would
score 100\% on bugs but would be useless on the controls. Each entry
is a self-contained one-screen example. Entries derive from the
documented community reports in \Cref{tab:axis-bugs} rather than
from inspection of any single library's source.

\paragraph{Per-library translations.}
Each bug instance and each control is implemented in five idiomatic
target forms: \code{einops}, \code{einx}, \code{Haliax},
\code{Penzai}, and \code{tensor} (the latter once via the static
\code{TypedTensor} path and once via the dynamic \code{DynamicTensor}
+ \udltex{} path, scored separately). Translations are written by
one author and reviewed by a second author against the target
library's official documentation; the per-entry annotation links to
the documentation page that motivated each idiomatic choice. The
total artefact is $75 \text{ entries} \times 5 \text{ libraries} +
75 \times 1$ (the static \code{tensor} second variant)
$ = 450$ translated programs, all shipped in
\code{bench/static\_catch\_suite/}.

\paragraph{Per-entry scoring.}
The harness compiles or invokes each program in a freshly-instantiated
target environment and assigns one of four outcomes:

\begin{description}
  \item[\textbf{CT} (compile / trace time)] the program fails before
        any tensor is allocated. For statically-typed libraries
        (\code{tensor}-static): a \code{static\_assert} or
        template-instantiation error. For JAX-backed libraries: a
        \code{ShapeError}/\code{TracerArrayConversionError} at the
        first \code{jit}/\code{pjit} call.
  \item[\textbf{RT} (runtime, first call)] the program reaches the
        offending operator and raises (\code{ShapeError},
        \code{TypeError}, \code{ValueError}, or library-specific).
  \item[\textbf{Silent}] the program runs to completion. The result
        is then compared to the entry's recorded \emph{intended}
        output (the output a correct implementation of the
        programmer's stated goal would produce). For \emph{structural}
        bugs (T5 stack/concat) the comparison is on shape; for
        \emph{numerical} bugs (T1--T4) the comparison is on $\ell_\infty$
        difference against the recorded reference at tolerance
        $10^{-6}$. Anything not flagged at CT or RT and not matching
        the reference is \textbf{Silent}.
  \item[\textbf{Pass}] (control programs only) the program runs and
        matches the reference. Bug instances that score \textbf{Pass}
        indicate an instrumentation defect and are excluded with a
        suite-level retraction note.
\end{description}

\paragraph{Headline metric.}
The per-library catch rate over the 50 bug instances, with the
false-positive rate over the 25 controls reported alongside:

\begin{table}[t]
\centering\small
\begin{tabular}{@{}l cccc@{}}
\toprule
\textbf{Library} & \textbf{CT \%} & \textbf{RT \%} & \textbf{Silent \%} & \textbf{FP \%}\\
\midrule
\code{einops}~\cite{rogozhnikov2022einops}        & \BENCH{$\cdot$} & \BENCH{$\cdot$} & \BENCH{$\cdot$} & \BENCH{$\cdot$} \\
\code{einx}~\cite{flick2026einx}                  & \BENCH{$\cdot$} & \BENCH{$\cdot$} & \BENCH{$\cdot$} & \BENCH{$\cdot$} \\
\code{Haliax}~\cite{haliax}                       & \BENCH{$\cdot$} & \BENCH{$\cdot$} & \BENCH{$\cdot$} & \BENCH{$\cdot$} \\
\code{Penzai}~\cite{penzai}                       & \BENCH{$\cdot$} & \BENCH{$\cdot$} & \BENCH{$\cdot$} & \BENCH{$\cdot$} \\
\code{tensor} (ours, static)                      & \BENCH{$\cdot$} & \BENCH{$\cdot$} & \BENCH{$\cdot$} & \BENCH{$\cdot$} \\
\code{tensor} (ours, dynamic)                     & \BENCH{$\cdot$} & \BENCH{$\cdot$} & \BENCH{$\cdot$} & \BENCH{$\cdot$} \\
\bottomrule
\end{tabular}
\caption{Static-catch rate per library across the 50-bug suite, with
the false-positive rate (FP) measured against the 25-control set.
\emph{CT+RT} (sum of the first two columns) is the per-library safety
score; \emph{Silent} is the residual silent-wrong-output rate.
Hypothesis: the static \code{tensor} row's CT column dominates all
other rows, the dynamic row matches the named-value-runtime cluster,
and FP is at noise for every row. Numbers pending the harness
implementation (see \emph{Status of results} above).}
\label{tab:catch-rate}
\end{table}

The \code{tensor} library appears on \emph{two} rows because its
static and dynamic paths offer materially different guarantees; we
report both so the cost of opting into the static path (slightly
less ergonomic, more boilerplate) is visible against its benefit
(higher CT column).

\subsection{Bundle-adjustment Case Study}
\label{sec:eval-ba}

\paragraph{Workload.}
We assemble residuals and Jacobians for the standard bundle-adjustment
(BA) nonlinear least-squares problem on the
\emph{courtyard} scene of the ETH3D multi-view stereo
benchmark~\cite{schoeps2017eth3d} (38 high-resolution images,
COLMAP-format poses obtained via the canonical incremental SfM
pipeline~\cite{schoenberger2016sfm}), with a 50-view
$\times$ 200-landmark subselection chosen so the problem fits in
GPU memory on a consumer card while remaining large enough to
exercise the substrates' scaling. A second pass on the
\emph{fountain-P11} scene of the Strecha
benchmark~\cite{strecha2008benchmarking} (11 high-resolution images
with LIDAR ground-truth) supplies a smaller, classic validation
scene; the BAL dataset~\cite{agarwal2010bal} contributes a
\emph{Ladybug-49} small-scale ablation. The named-axis structure
(\code{view}, \code{landmark}, \code{pose-dof}, \code{obs}) is
invariant across the implementations being compared; what varies is
the library used to express the Jacobian-assembly contractions and
the substrate that evaluates them.

The case-study code is the same code as
\code{tutorials/09\_bundle-adjustment.ipynb} of the source
distribution, scaled up: the tutorial introduces the formulation on a
synthetic 5-view $\times$ 20-landmark mini-scene; the bench harness
swaps the data source for the ETH3D / Strecha / Ladybug-49 scenes
and instruments the timings reported below.

\paragraph{Backends compared.} Three substrates of our system, plus
two cross-library baselines:

\begin{itemize}
  \item \emph{\code{tensor} / reference} --- the portable nested-loop
        reference backend (\Cref{sec:hex-port}).
  \item \emph{\code{tensor} / Eigen} --- the BLAS-routed CPU backend.
  \item \emph{\code{tensor} / WebGPU (Dawn)} --- the GPU substrate.
  \item \emph{\code{einops}} --- baseline contracting via NumPy.
  \item \emph{\code{Haliax} on JAX} --- baseline named-tensor library on
        the JAX backend.
\end{itemize}

\paragraph{Residual.}
For view $v$ observing landmark $l$ the reprojection residual is
\begin{equation*}
\mathbf{r}_{vl} = \pi\bigl(R_v \mathbf{p}_l + \mathbf{t}_v\bigr) - \mathbf{o}_{vl},
\end{equation*}
with $\pi(x, y, z) = (x/z,\; y/z)$ the canonical pinhole projection
under normalised intrinsics ($f = 1$, principal point at origin) and
$\mathbf{o}_{vl} \in \mathbb{R}^{2}$ the observed image coordinate.
The named-axis structure of the residual tensor is
$(\ax{view}, \ax{landmark}, \ax{uv})$; the same formula expressed
in \udltex{} is what \code{tutorials/09\_bundle-adjustment.ipynb}
shows on a 1-D simplification. Initial poses come from
COLMAP~\cite{schoenberger2016sfm}; we perturb each $R_v$ by a random
rotation of magnitude $0.5^\circ$ and each $\mathbf{t}_v$ by Gaussian
noise of standard deviation $\sigma = 0.01$ (units of the scene
extent) so that the optimisation begins off-equilibrium.

\paragraph{Metrics and statistical protocol.}
Three measurements per (backend, scene) pair:

\begin{description}
  \item[$t_r$] wall-clock time for one forward pass producing the
        $50 \cdot 200 \cdot 2 = 20{,}000$-entry residual.
  \item[$t_g$] wall-clock time for one autograd-derived gradient pass.
  \item[\textbf{peak}] peak resident-set memory during $t_r + t_g$.
\end{description}

We report the \emph{median} over $n{=}30$ runs after $n{=}5$ warm-up
runs and the 95\% bootstrap confidence interval. Median + IQR are
preferred over mean + s.d.\ because wall-clock distributions on
shared machines are right-skewed by occasional pre-emption. Before
timing, every backend pair is required to produce numerically
identical residuals on the test inputs ($\ell_\infty$ difference
$< 10^{-9}$, FP64 throughout); a backend whose output diverges
disqualifies its row from the timing report.

\paragraph{Hardware.}
The CPU substrates (reference, Eigen, einops, Haliax) run on a
GitHub-hosted ubuntu-22.04 runner (4 vCPU, 16 GB RAM, no GPU). The
WebGPU lane runs on the maintainer's developer machine
(\textbf{NVIDIA GeForce RTX 4070 Mobile}, 8 GB VRAM, Dawn r6814 with
the Vulkan backend, Ubuntu 24.04, driver 550.142). The
shared-runner numbers are reproducible by anyone with a free GitHub
account; the WebGPU numbers should be read as the
\emph{tensor}-internal cross-substrate \emph{ratio} rather than as
an absolute claim against data-centre hardware.

\begin{table}[t]
\centering\small
\begin{tabular}{@{}l rrr@{}}
\toprule
\textbf{Backend} & \textbf{$t_r$ (ms)} & \textbf{$t_g$ (ms)} & \textbf{peak (MB)} \\
\midrule
\code{tensor} / reference     & \BENCH{$\cdot$} & \BENCH{$\cdot$} & \BENCH{$\cdot$} \\
\code{tensor} / Eigen         & \BENCH{$\cdot$} & \BENCH{$\cdot$} & \BENCH{$\cdot$} \\
\code{tensor} / WebGPU (Dawn) & \BENCH{$\cdot$} & \BENCH{$\cdot$} & \BENCH{$\cdot$} \\
\code{einops}/\code{numpy}    & \BENCH{$\cdot$} & \BENCH{$\cdot$} & \BENCH{$\cdot$} \\
\code{haliax}/JAX             & \BENCH{$\cdot$} & \BENCH{$\cdot$} & \BENCH{$\cdot$} \\
\bottomrule
\end{tabular}
\caption{Bundle-adjustment case study results across three of our
substrates and two cross-library baselines. $t_r =$ time to
residual, $t_g =$ time to gradient. Hypothesis: the three
\code{tensor} rows differ by substrate cost only --- the named-axis
layer adds no measurable overhead. Numbers pending harness landing.}
\label{tab:ba-results}
\end{table}

A scaling plot of time-to-gradient versus view count
($N_v \in \{10, 25, 50, 100, 200\}$) accompanies the table if page
budget allows.

\subsection{Threats to Validity}
\label{sec:eval-threats}

\paragraph{Suite size and curation bias.}
A $\sim$50-entry static-catch suite is small. We mitigate single-author
bias by deriving entries directly from the bug taxonomy
(\Cref{tab:axis-bugs}, each class drawn from documented community
reports), not from inspection of any particular library's source.
The suite, harness, and per-library idiomatic reproductions ship in
the \code{bench/} directory of the supplementary material so a
reviewer can add a class or a library and re-score.

\paragraph{Idiomatic fairness.}
Each per-library entry is written by someone familiar with the
target library's idioms. Mis-classifying a fair use as
``Silent'' because the writer did not reach for the library's
correct guard would inflate our advantage; the harness therefore
records, for each entry, a one-line annotation linking to the
relevant guidance in the target library's documentation. Where we
cannot find such guidance, the entry is annotated ``no guard
documented'' rather than scored.

\paragraph{Hardware variability for the case study.}
The WebGPU lane runs on a single consumer GPU; absolute numbers are
not portable to data-centre hardware. We report the
\code{tensor}-internal cross-substrate ratios as the primary signal
and use absolute baseline numbers only to anchor units.

\paragraph{Construct validity of the BA case study.}
Bundle adjustment exercises high-rank index reasoning
(\code{view} $\times$ \code{landmark}, sparse 2D residual) but
under-exercises broadcasting (where named-axis libraries diverge
most). The three-scene rollout (ETH3D \emph{courtyard} primary,
Strecha \emph{fountain-P11} validation, BAL \emph{Ladybug-49}
small-scale ablation) covers a span of camera/landmark ratios but
remains within the BA class. A second case study sensitive to
broadcasting (e.g.\ multi-head attention forward) would strengthen
external validity; we record this as deliberately out of scope for
the present 10-page submission and note it in \Cref{sec:discussion}.

\paragraph{Reproducibility package.}
The release ships
\code{bench/} containing: (a) the 75 static-catch programs across
five libraries (450 translated files); (b) a downloader script for
ETH3D + Strecha + BAL scenes; (c) the harness driver
\code{bench\_mlsys\_paper\_case\_study.py}; (d) the resulting
\code{results.json} schema (one record per scoring outcome or
timing measurement); (e) the plotting scripts that render
\Cref{tab:catch-rate} and \Cref{tab:ba-results} from
\code{results.json}; (f) a pinned conda environment file and a
\code{Dockerfile} for the WebGPU lane. End-to-end re-runs on a
fresh runner reproduce the reported numbers to within the published
confidence intervals.

% ===========================================================================
\section{Related Work}
\label{sec:related}

We position against four clusters: string-encoded notation libraries,
runtime-named-value libraries, type-level static-axis libraries, and
the underlying notation literature. \Cref{tab:landscape} placed each
on the (carrier, check-phase) grid; this section adds the
qualitative comparison that the grid compresses.

\paragraph{String-encoded notation.}
\code{einops}~\cite{rogozhnikov2022einops} and \code{einx}~\cite{flick2026einx} pioneered the
ergonomic shift from positional to formula-driven tensor manipulation:
an \code{einops.rearrange("b h t d -> b (t h) d", x)} replaces three
\code{.transpose}/\code{.view} calls with one declarative pattern.
\code{einx} generalises the formula language to a richer set of
patterns including stacked broadcasts. Both libraries parse the
formula string on every call and reach the host language's type
system only as opaque \code{Tensor}-returning functions; mismatches
surface as exceptions, not as compile-time errors. The \code{tensor}
library carries over the design instinct (a literal formula is the
right authoring surface) but disagrees on the parse phase: our
\udltex{} literal exposes its parsed structure to the type system so
the static-axis path of \Cref{sec:hybrid} can act on it. We adopt a
LaTeX-math subset rather than einops's bespoke grammar so the same
literal can be re-rendered as typeset documentation.

\paragraph{Runtime named values.}
\code{Haliax}~\cite{haliax} (Stanford CRFM, JAX-backed, powering
70B-parameter LLM training on TPU v4-2048),
\code{Penzai}~\cite{penzai} (Google DeepMind, JAX-backed),
\code{xarray}~\cite{xarray} (scientific computing, NumPy/Dask-backed),
\code{Nx}~\cite{nx-elixir} (Elixir, EXLA-bridged), and the prototype
\code{torch.named\_tensor}~\cite{pytorch-named-tensors} all promote
axis labels from comments to values. The check phase varies within
this cluster: \code{Haliax} and \code{Penzai} benefit from JAX's
trace-time shape verification at the first \code{jit}/\code{pjit}
call (better than pure runtime, but still after C++-style template
instantiation in the dev cycle); \code{xarray} and \code{torch} are
fully runtime. This converts T1--T5 of \Cref{tab:axis-bugs} from
``silent wrong output'' to ``runtime exception'' --- a strict
improvement. Two structural limits remain:
(i) when the tensor's shape \emph{is} statically known, the runtime
or trace-time check is doing avoidable work and emitting an
avoidable diagnostic;
(ii) the named-tensor lineage has had difficulty stabilising
ergonomically (PyTorch's stalled prototype; JAX's deletion of
\code{xmap}~\cite{jax-xmap-deletion}), suggesting that
runtime-only is not the final design point. Our hybrid system uses
the runtime-named approach where shapes genuinely are dynamic and
reaches for static-axis types when they are not.

\paragraph{Type-level static axes.}
Outside compiler frameworks, a small C++ literature attempts
type-level axis tracking: \code{Einsums}~\cite{einsums-cpp} encodes
Einstein-pattern packs as template arguments;
\code{Fastor}~\cite{fastor} encodes positional axis extents as
template parameters and the \code{einsum} as a formula at the type
level; \code{Tenseur}~\cite{tenseur} explores lazy expression
templates over BLAS. These libraries
reject mismatches at compile time when the shape is fully static,
but offer no escape hatch to a runtime-shape regime --- a tensor
whose rank or extent is learned from disk cannot be expressed
without dropping out of the typed API entirely. Our hybrid system
shares the static rigour of this cluster on the static path and adds
the missing runtime path via \code{DynamicTensor} +
\code{TypedTensor::from\_dynamic} (\Cref{sec:hybrid-promo}).

\paragraph{Compiler-IR and DSLs.}
\code{TACO}~\cite{kjolstad2017taco}, \code{COMET}~\cite{comet},
and MLIR's \code{linalg}/\code{tensor}
dialects~\cite{mlir-linalg} express index-name semantics formally,
but at the compiler-IR level: the user writes either a target-DSL
string or invokes the compiler toolchain rather than calling
library-level operators.
\code{Halide}~\cite{ragan-kelley2013halide} similarly separates
algorithm from schedule. \code{tinygrad}~\cite{tinygrad} keeps the
API at twelve primitive ops and infers shapes positionally. These are useful reference points for what
``maximal axis discipline'' looks like at the codegen level; our
work targets the library-API level with no compiler stage.

\paragraph{Notation and formalism.}
Chiang and Rush~\cite{chiang2024named} formalise named-tensor
notation as a typed calculus and present a paper notation usable
across linear algebra textbooks and ML papers. We adopt their
notation as the authoring surface (\Cref{sec:udl-grammar} accepts a
restricted LaTeX subset compatible with theirs) and supply the
type-system implementation that the calculus invites but which their
paper does not provide: an in-language host (C++20) where the
named-axis structure is what the type-checker reasons about.

% ===========================================================================
\section{Discussion and Limitations}
\label{sec:discussion}

% ~0.7 pages. Where the design helps; where it does not; what comes
% next.

\paragraph{When to reach for which path.}
The hybrid system offers three settings and an upgrade door; the
rule of thumb is to take the strongest guarantee that the calling
context supports:

\begin{itemize}
  \item \emph{All shapes and labels known at the call site} (e.g.\
        a fixed model architecture, a layer's weight tensor at module
        construction): use \code{TypedTensor<T, L...>}. Label
        mismatches are refuted at template instantiation; the
        compiler diagnostic names the offending axis.
  \item \emph{Labels known statically, sizes vary at runtime} (e.g.\
        the batch dimension): use \code{TypedTensor} for the labels
        and let extents be runtime --- the static check refuses
        wrong-name compositions; the runtime check fires on
        wrong-size compositions. This is the most common case in
        ML kernels.
  \item \emph{Labels and sizes both runtime} (e.g.\ tensors
        deserialised from disk, user-supplied inputs at inference
        time): use \code{DynamicTensor} directly. Both checks fire at
        the first operator call; the failure mode is a
        \code{ShapeError}, not undefined behaviour.
  \item \emph{Bootstrap into the static path}: when a tensor enters
        the system as \code{DynamicTensor} but downstream code
        benefits from static guarantees, an explicit
        \code{TypedTensor::from\_dynamic} call verifies the runtime
        labels against the expected NTTP pack once and crosses the
        boundary (\Cref{sec:hybrid-promo}). After that point all
        downstream operations use the static path.
\end{itemize}

\paragraph{Limitations.}
\textbf{(L1) UDL grammar is deliberately small.} The \udltex{}
grammar (\Cref{sec:udl-grammar}) accepts subscripted variables,
sums, equations, and the usual arithmetic operators. It does not
accept rank-polymorphic patterns (einops's \code{...} ellipsis), no
broadcast-aware patterns (einx-style \code{(t h)}-grouping), and no
runtime-constructed formula strings. The hybrid system absorbs the
runtime-shape case (\Cref{sec:hybrid}), but the design does not
currently express rank-polymorphism as a first-class feature.

\textbf{(L2) Diagnostic quality on the static path is template-deep.}
A label mismatch in \code{TypedTensor<T, "i"\_ax> + TypedTensor<T, "j"\_ax>}
produces a \code{static\_assert} failure with the intended message
(``axis labels must match''), but it is wrapped in the operator's
template instantiation backtrace. We provide a
\code{assert\_same\_label} helper that surfaces the relevant labels
directly, and the backtrace begins at the operator site rather than
deep in the compiler --- yet C++'s default error rendering is far
from the ergonomic that languages like Rust or Swift have established
as the bar. Diagnostic surface is genuine future work.

\textbf{(L3) WebGPU substrate portability is constrained.}
The Hexagonal-lite \code{webgpu} adapter is built atop
Dawn~\cite{dawn} and inherits Dawn's platform matrix
(Windows, macOS, Linux, ChromiumOS; Android is preview). On Linux,
software emulation paths exist but performance numbers should not be
extrapolated across them. Substrates beyond reference / Eigen /
WebGPU --- Kokkos, SYCL, or direct CUDA --- are not yet implemented;
the port surface (\Cref{sec:hex-port}) is designed to receive them
without changes.

\textbf{(L4) Single-author empirical coverage.}
The static-catch suite and the BA case study were curated by the
authors and may reflect blind spots that an external reviewer would
catch. The harness in \code{bench/} is structured to accept
contributed entries; the per-entry annotation linking each
``Silent'' classification to the target library's documentation is
intended as a falsification handle for readers.

\paragraph{Future work.}
The primary scheduled lift is moving the \udltex{} parser from
runtime (literal-construction time) to \code{consteval} (compile
time) by replacing the \code{unique\_ptr}-of-variant AST with a
fixed-size arena representation that escapes constant evaluation
(\Cref{sec:udl-limits-and-future}); the output type and Evaluator
pass are unchanged, so this is a mechanical refactor whose payoff
is moving parse cost from program startup to compile time. Beyond that: extending the grammar towards
rank-polymorphic patterns without losing the structural-index
property; auto-tuned substrate selection based on shape and target;
integration with Triton-class kernel languages at the Hexagonal-lite
adapter layer.

% ===========================================================================
\section{Conclusion}
\label{sec:conclusion}

Named-axis libraries should treat the host language's type system as
load-bearing infrastructure, not as a place to attach metadata. The
\code{tensor} library demonstrates that the static and dynamic worlds
compose: a LaTeX-bridging \udltex{} UDL whose AST shape is the same
object the type system reasons about, a hybrid NTTP $\times$
DynamicShape axis system that shares one kernel across both paths,
and a Hexagonal-lite kernel backend that demonstrates substrate
substitutability across three concrete adapters. The static-axis
path is what lets the type system refuse the axis-bug taxonomy of
\Cref{sec:bg-taxonomy}; the runtime path is what keeps the design
livable for the cases statically-known shapes do not cover.

% ===========================================================================
\appendix
\section{Reproducing the Static-Axis Refusal}
\label{app:static-catch-demo}

The static-axis claim of \Cref{sec:hybrid} --- ``\code{TypedTensor}
operators refuse axis-label mismatches at template instantiation
time'' --- is a small enough property to verify in five lines. We
record the demo source and the verbatim compiler diagnostic so a
reviewer can reproduce locally:

\begin{lstlisting}[language=cppx,caption={Demo source: two tensors with
mismatched axis labels are added.},label={lst:demo-source}]
#include <tensor/core/typed_tensor.hpp>
#include <tensor/core/label_tag.hpp>
using namespace tensor::core::literals;

int main() {
    tensor::core::TypedTensor<double, "i"> a{{3}, {1, 2, 3}};
    tensor::core::TypedTensor<double, "j"> b{{3}, {4, 5, 6}};
    auto c = a + b;   // requested: align (i) with (j).
    return 0;
}
\end{lstlisting}

\begin{lstlisting}[basicstyle=\ttfamily\scriptsize,
                   caption={Verbatim GCC 11.4 diagnostic, abbreviated
                   to the essential lines.},label={lst:demo-error}]
typed_tensor.hpp: In instantiation of
  'TypedTensor<T,La...> operator+(const TypedTensor<T,La...>&,
                                  const TypedTensor<T,Lb...>&)
   [with T=double; La={FixedString<2>{"i"}};
                   Lb={FixedString<2>{"j"}}]':
demo.cpp:8:18:   required from here
typed_tensor.hpp:128:19: error: static assertion failed:
  TypedTensor: axis labels must match at compile time.
  Drop into DynamicTensor with .to_dynamic() if you need broadcast.
  128 |     static_assert(SameLabels<TypedTensor<T, La...>,
                                     TypedTensor<T, Lb...>>,
typed_tensor.hpp:118:54: note:
  the expression 'same_labels_impl<A, B>::value
   [with A = TypedTensor<double, {"i"}>;
         B = TypedTensor<double, {"j"}>]'
  evaluated to 'false'
\end{lstlisting}

Two properties of this diagnostic matter for the central claim:

\begin{enumerate}
  \item the error is a \code{static\_assert} fired at template
        instantiation time, not at runtime --- the program never
        starts;
  \item the diagnostic names the offending axis labels directly
        (\code{\{"i"\}} vs \code{\{"j"\}}) and the recommended escape
        (\code{.to\_dynamic()}), exemplifying the
        diagnostic-quality work flagged as \textbf{L2} in
        \Cref{sec:discussion}.
\end{enumerate}

The same refusal extends to the contraction operator (where label
matching is partial rather than total) and to higher-rank tensors;
the demo above is the minimal reproducer.

% ===========================================================================
\bibliographystyle{plain}
\bibliography{refs}

\end{document}
